---
engine: knitr
execute:
  eval: false
---

# RStudio Getting Started

## What is RStudio?

**Positron is the recommended IDE for USACE data science** when you are working on R and Python based technical analysis projects. **RStudio Desktop** remains a strong, widely used IDE built specifically for the **R** language, and it is a good choice for USACE work that is primarily R-based—especially when your workflow emphasizes:

- R scripts and packages (including package development),
- reproducible reports (R Markdown / Quarto),
- interactive exploration (console, plots, environment, viewer panes),
- a stable, purpose-built R experience with a large existing user base.

This chapter is intended to **aim you in the right direction** (USACE install path + prerequisites + basic verification). It does not duplicate Posit’s authoritative documentation.

Authoritative documentation:
- RStudio (Posit) documentation / [RStudio landing page](https://posit.co/products/open-source/rstudio/)
- Posit Support (general): [support.posit.co/](https://support.posit.co/)

## When should I use RStudio vs Positron?

As a rule for USACE technical analysis projects:

- Choose **Positron** for most new work, especially when your projects span **R + Python** or you want a more VS Code-like experience.
- Choose **RStudio** when your work is **R-first** and you want the most established, R-specific IDE experience.
- If your team standardizes on one IDE for supportability, follow team guidance.

## Prerequisites (install these first)

Before installing or using RStudio for R-based work, make sure your system has:

1. **R** (required)
2. **RTools** (Windows only; sometimes required for building packages from source)

### Install R (required)

R provides the runtime that RStudio uses to run your scripts and render R/knitr outputs.

- R homepage: [https://www.r-project.org/](https://www.r-project.org/)

USACE users should follow the **approved/scanned software** process described in the “Open Source Software” chapter:

- See: [Open Source Software](gov-open-source-software.qmd) (Status Check + App Portal)

### Install RTools (Windows; “maybe” required)

You may need **RTools** on Windows when installing R packages that compile from source (C/C++/Fortran) or when your workflow requires building packages locally.

- RTools: [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/)

If you are unsure: install R first, then attempt to install your required packages. If package installation fails with compilation/toolchain errors, install RTools and retry.

## Install and configure RStudio (USACE pathway)

::: {.callout-note}
## USACE installation path (App Portal)

For USACE-managed devices, install software through the USACE **App Portal** and confirm you are using an approved/scanned version.

- Check scanned/approved versions: [Finder](https://finder.nosc.usace.army.mil/code_scan_tracker)
- Install approved versions via the USACE App Portal: [App Portal](https://app-portal.usace.army.mil/)

These links and the “Status Check” workflow are maintained in: [Open Source Software](gov-open-source-software.qmd).
:::

After RStudio is installed, do a quick verification to ensure RStudio can run R successfully.

### First launch checklist (R / knitr projects)

1. Open RStudio.
2. Confirm RStudio is using the expected R installation:
   - Tools → Global Options → General → Basic → “R version”
3. Open a new R script (for example: `scratch.R`).
4. Run the following in the R console to confirm R is available:

```{r}
version
```

You should see your installed R version printed.

If you plan to render Quarto/knitr documents locally, also confirm you can install and load common packages (your project may require additional packages):

```{r}
install.packages(c("knitr", "rmarkdown", "quarto"))
```

If installs fail with compilation errors on Windows, revisit the RTools prerequisite.

## Recommended baseline settings (team-friendly defaults)

These are optional, but commonly help with consistency and reproducibility across a team:

- Prefer working in a dedicated **Project** (File → New Project) rather than ad-hoc scripts scattered across the filesystem.
- Keep analysis dependencies explicit:
  - consider an R package lockfile workflow (e.g., `renv`) when projects need to be re-run reliably on other machines.
- Avoid relying on hidden local state:
  - restart your R session periodically,
  - re-run scripts from a clean session to confirm they work end-to-end.

(Choose conventions deliberately and document them in your project README.)

## Getting help

If something fails (install, packages, R version detection), capture:

- exact error message,
- R version (`version`),
- RStudio version (Help → About),
- whether you are on a USACE-managed device (App Portal-installed).

Then consult:
- Posit’s [RStudio documentation](https://posit.co/products/open-source/rstudio/)
